Predicting the effect of candidate interventions for mitigating disease spread in asymmetric-mixing metapopulations

### 2024-07-26 ms.tex
going to write analysis sketch
added thought about equity and planning to disc

Despite some evidence to support the hypothesis that young male
Indigenous mobility can explain transmission into the reserve, there are
many complicating factors that make it difficult to evaluate this
hypothesis directly without adding some additional qualifications. 
There
may be other social factors that may be necessary for the time series
observations of De Oliveira and co-authors. 
Furthermore, there may be
less significant factors driving transmission in this model that may be
of greater importance in future outbreaks of different diseases than
SARS-CoV-2 in the reserve. 
Therefore, here we expand the scope of the
hypothesis to include hypotheses about how social groups, namely village
residency and Indigenous ethnic group, affect transmission in the
reserve and explain observed outbreaks of variants B.1.1 and Zeta in the
Reserve. To evaluate the hypotheses, presented in the next paragraph, we
develop a mechanistic model of epidemiological transmission with
asymmetric social mixing that evaluates how best to model group
structure in the Reserve. In turn, this model is used to evaluate
potential economic interventions that target aid to different groups to
reduce infections in the reserve.

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:Map} ABOUT HERE ]} \\
\end{center}

@@

Our model represents interactions within and between Dourados Indigenous 
Reserve groups, and with members of the city of Dourados 
(Figure~\ref{fig:ModelSketch}). To represent
interactions between groups we define two possible groupings, by village
(Figure~\ref{fig:ModelSketch}, Model 1) and by ethnicity
(Figure~\ref{fig:ModelSketch}, Model 2). The within-group contact rate 
is $c_{gg}$ for focal group $g$. The contact rate between the focal group and
its outgroup, $g'$, is $c_{gg'}$. If in Model 2 the focal group is Terena, then
the within-group contact rate is $c_{TT}$ and the between-group contact rate,
representing how frequently Terena interact with Guarani is $c_{TG}$, Each of these groupings can be further
subdivided by the alternative Model grouping, represented in 
Figure~\ref{fig:ModelSketch} by the inset pie chart diagrams in each group's
circle. Reserve resident mobility is represented by a contact-infection rate,
$\gamma_g$, which represents how frequently a susceptible member of group $g$
brings an infection into the Reserve. The exact compartments, $S$, $I$, and $R$
for the city population are ignored.

@@

% \vspace{0.5em}
% \begin{center}
% {[ FIGURE \ref{fig:ByReinfectionByStrategy} ABOUT HERE ]} \\
% \end{center}

@@

\begin{itemize}
  \item 
    Application to other problematic diseases in the Reserve, e.g., TB, STDs,\ldots 
  \item
    The data is not necessarily representative and we had to use educated guesses
    for mixing parameters.
  \item
    Have not yet considered more complex epidemiological and behavioral
    dynamics. For example, group structure can affect the coupled contagion of
    disease and behaviors that could encourage or impede the spread of
    disease~\autocite{SmaldinoJones2021}. Our compartmental model
    incorporated group structure and contact rate heterogeneity, but finer
    details of the social network structure can induce complex
    epidemiological dynamics~\autocite{Arthur2017}. In the case of Dourados
    and possibly other Indenous communities is the tendency for several
    generations of families to live in close proximinty to one another,
    often communally with other families as well, making social isolation
    culturally unpalatable and, consequentially, unfeasible. Adaptation to
    evolving, uncertain epidemiological social and environmental contexts is
    one further factor that can complicate epidemiological
    dynamics~\autocite{Arthur2021} and willingness to learn beneficial public
    health behaviors from others~\autocite{Turner2023a}.
\end{itemize}

@@

\begin{figure}[H]
\caption{Time series of COVID-19 spread through the Dourados Indigenous
Reserve, with significant events highlighted. The first B.1.1 variant
cases were introduced by the Guarani ethnicity (1), followed about one
month later by a rapid rise in cases in Jaguapiru among the Terena
ethnicity (2); several months later a rapid rise in Bororó was caused by
a rapid rise among Guarani (3). Zeta was introduced at the end of 2020
(4), with a rapid rise around the holiday season especially in Jaguapiru
among Guarani (5); later in February cases rose in Bororó among the
Terena (6).}
\label{fig:DeOliveiraSeries}
\vspace{0.3em}
{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{Figures/Time Series.pdf}

@@

\caption{
  We can model the relevant groups for mixing as either village
residence or ethnicity. Mixing, and mobility parameters
would be set accordingly. Mixing is asymmetric, with $c_{gg}$ representing the
contact rate within group $g$, and each $c_{gg'}$ representing
the contact rate at which a member of group $g$ encounters a member of the
outgroup $g'$, e.g., $c_{GT}$ is the contact rate at which Guarani encounter
members of the Terena. The $\gamma_g$ parameters would be reduced through
economic aid to group $g$ for them to remain on the Reserve during
outbreaks. \mt{Looks like some labels are cut off; will fix in later draft.}}
\label{fig:ModelSketch}
\vspace{0.3em}
{\centering \includegraphics{Figures/ModelSketch.pdf}

>>>>

Despite some evidence to support the hypothesis that young male
Indigenous mobility can explain transmission into the reserve, there are
many complicating factors that make it difficult to evaluate this
hypothesis directly without adding some additional qualifications. 
There
may be other social factors that may be necessary for the time series
observations of De Oliveira and co-authors. 
Furthermore, there may be
less significant factors driving transmission in this model that may be
of greater importance in future outbreaks of different diseases than
SARS-CoV-2 in the reserve. 
Therefore, here we expand the scope of the
hypothesis to include hypotheses about how social groups, namely village
residency and Indigenous ethnic group, affect transmission in the
reserve and explain observed outbreaks of variants B.1.1 and Zeta in the
Reserve. To evaluate the hypotheses, presented in the next paragraph, we
develop a mechanistic model of epidemiological transmission with
asymmetric social mixing that evaluates how best to model group
structure in the Reserve. In turn, this model is used to evaluate
potential economic interventions that target aid to different groups to
reduce infections in the reserve. The effect of different amounts of aid to
different groups 
\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:Map} ABOUT HERE ]} \\
\end{center}

@@

Our model represents interactions within and between Dourados Indigenous 
Reserve groups, and with members of the city of Dourados 
(Figure~\ref{fig:ModelSketch}). To represent
interactions between groups we define two possible groupings, by village
(Figure~\ref{fig:ModelSketch}, Model 1) and by ethnicity
(Figure~\ref{fig:ModelSketch}, Model 2). The within-group contact rate 
is $c_{gg}$ for focal group $g$. The contact rate between the focal group and
its outgroup, $g'$, is $c_{gg'}$. If in Model 2 the focal group is Terena, then
the within-group contact rate is $c_{TT}$ and the between-group contact rate,
representing how frequently Terena interact with Guarani is $c_{TG}$, Each of these groupings can be further
subdivided by the alternative Model grouping, represented in 
Figure~\ref{fig:ModelSketch} by the inset pie chart diagrams in each group's
circle. Reserve resident mobility is represented by a contact-infection rate,
$\gamma_g$, which represents how frequently a susceptible member of group $g$
brings an infection into the Reserve. The exact compartments, $S$, $I$, and $R$
for the city population are ignored.
The $\gamma_g$ parameters are proxies for resources allocated to 
economic aid or other programs to support to group $g$ to remain on the Reserve 
during outbreaks.

@@

@@

\begin{itemize}
  \item 
    Application to other problematic diseases in the Reserve, e.g., TB, STDs,\ldots 
  \item
    The data is not necessarily representative and we had to use educated guesses
    for mixing parameters.
  \item
    Have not yet considered more complex epidemiological and behavioral
    dynamics. For example, group structure can affect the coupled contagion of
    disease and behaviors that could encourage or impede the spread of
    disease~\autocite{SmaldinoJones2021}. Our compartmental model
    incorporated group structure and contact rate heterogeneity, but finer
    details of the social network structure can induce complex
    epidemiological dynamics~\autocite{Arthur2017}. In the case of Dourados
    and possibly other Indenous communities is the tendency for several
    generations of families to live in close proximinty to one another,
    often communally with other families as well, making social isolation
    culturally unpalatable and, consequentially, unfeasible. Adaptation to
    evolving, uncertain epidemiological social and environmental contexts is
    one further factor that can complicate epidemiological
    dynamics~\autocite{Arthur2021} and willingness to learn beneficial public
    health behaviors from others~\autocite{Turner2023a}.
  \item
    Currently we do not connect how much it costs to implement these programs,
    and there may be difficult questions over what equality means: is it
    spending the same amount of money per group or preventing the same number of
    hypothetical infections for each group? Of course a scientific article 
    cannot provide the answer, but here we do provide a program for evaluating
    different choices to prototype planned responses to pandemic outbreaks and
    reason quantitatively about alternative courses of action.
\end{itemize}

@@

\begin{figure}[H]
\caption{\textbf{Time series of COVID-19 spread through the Dourados Indigenous
Reserve,} with significant events highlighted. The first B.1.1 variant
cases were introduced by the Guarani ethnicity (1), followed about one
month later by a rapid rise in cases in Jaguapiru among the Terena
ethnicity (2); several months later a rapid rise in Bororó was caused by
a rapid rise among Guarani (3). Zeta was introduced at the end of 2020
(4), with a rapid rise around the holiday season especially in Jaguapiru
among Guarani (5); later in February cases rose in Bororó among the
Terena (6).}
\label{fig:DeOliveiraSeries}
\vspace{0.3em}
{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{Figures/Time Series.pdf}

@@

\caption{
\textbf{Relevant groups for modeling inter-group mixing could be village
residence or ethnicity.} Mixing is asymmetric, with $c_{gg}$ representing the
contact rate within group $g$, and each $c_{gg'}$ representing
the contact rate at which a member of group $g$ encounters a member of the
outgroup $g'$, e.g., $c_{GT}$ is the contact rate at which Guarani encounter
members of the Terena (or vice-versa). \mt{Looks like some labels are cut off; will fix in later draft.}}
\label{fig:ModelSketch}
\vspace{0.3em}
{\centering \includegraphics{Figures/ModelSketch.pdf}
